# Example datasets

## CRC training data

This dataset comes from the `r Biocexptpkg("MSstatsBioData")` package
and was generated as follows:

```{r crctrain, eval = FALSE}
library("MSstats")
library("MSstatsBioData")
data(SRM_crc_training)
Quant <- dataProcess(SRM_crc_training)
subjectQuant <- quantification(Quant)
```

which provides quantitative information for 72 proteins, including two
standard proteins, *AIAG-Bovine* and *FETUA-Bovine*. These proteins
were targeted for plasma samples with SRM with isotope labeled
reference peptides in order to identify candidate protein biomarker
for non-invasive detection of CRC. The training cohort included 100
subjects in control group and 100 subjects with CRC. Each sample for
subject was measured in a single injection without technical
replicate. The training cohort was analyzed with Skyline. The dataset
was already normalized as described in manuscript. User do not need
extra normalization. NAs should be considered as censored missing. Two
standard proteins can be removed for statistical analysis.


Clinical information where added manually thereafter.

To load this dataset:

```{r crcread}
crcdf <- read.csv("http://bit.ly/VisBiomedDataCrcCsv")
crcdf[1:10, 1:3]
crcdf[1:10, 73:79]
```

Conversion to an `MSnSet`

```{r crcmsnset}
library("MSnbase")
i <- 1:72
e <- t(crcdf[, i])
colnames(e) <- 1:200
crc <- readMSnSet2(data.frame(e), e = 1:200)

pd <- crcdf[, -i]
rownames(pd) <- paste0("X", rownames(pd))
pData(crc) <- pd
crc
```


Or load it direcly:

```{r crcload}
download.file("http://bit.ly/VisBiomedDataCrcMSnSet", "./data/crc.rda")
load("./data/crc.rda")
crc
```

Reference:

See Surinova, S. et al. (2015) *Prediction of colorectal cancer
diagnosis based on circulating plasma proteins*. EMBO Mol. Med., 7,
1166–1178 for details.

## ALL data

```{r all}
library("ALL")
data(ALL)
ALL
```

From the documentation page:

> The *Acute Lymphoblastic Leukemia Data from the Ritz Laboratory*
> consist of microarrays from 128 different individuals with acute
> lymphoblastic leukemia (ALL). A number of additional covariates are
> available. The data have been normalized (using rma) and it is the
> jointly normalized data that are available here.

Reference:

Sabina Chiaretti, Xiaochun Li, Robert Gentleman, Antonella Vitale,
Marco Vignetti, Franco Mandelli, Jerome Ritz, and Robin Foa *Gene
expression profile of adult T-cell acute lymphocytic leukemia
identifies distinct subsets of patients with different response to
therapy and survival*. Blood, 1 April 2004, Vol. 103, No. 7.
