# Plotting in R

## Base graphics

We won't go through all base graphics plotting functions one by one
here. We will encounter and learn several of these functions
throughout the course and, if necessary, discuss them when questions
arise.

## Plotting with `ggplot2`

Base graphics uses a *canvas model* a series of instructions that
sequentially fill the plotting canvas. While this model is very useful
to build plots bits by bits bottom up, which is useful in some cases,
it has some clear drawback:

* Layout choices have to be made without global overview over what may
  still be coming.
* Different functions for different plot types with different
  interfaces.
* No standard data input.
* Many routine tasks require a lot of boilerplate code.
* No concept of facets/lattices/viewports.
* Poor default colours.

The `ggplot2` package implements a grammar of graphics. Users describe
what and how to visualise data and the package then generates the
figure. The components of `ggplot2`'s of graphics are

1. A **tidy** dataset
2. A choice of geometric objects that servers as the visual
   representation of the data - for instance, points, lines,
   rectangles, contours.
3. A description of how the variables in the data are mapped to visual
   properties (aesthetics) or the geometric objects, and an associated
   scale (e.g. linear, logarithmic, polar)
4. A statistical summarisation rule
5. A coordinate system.
6. A facet specification, i.e. the use of several plots to look at the
   same data.



Fist of all, we need to load the `ggplot2` package and load the `iprg` data.

```{r iprg}
library("ggplot2")
iprg <- read.csv("http://bit.ly/VisBiomedDataIprgCsv")
```

ggplot graphics are built step by step by adding new elements.

To build a ggplot we need to:

* bind the plot to a specific data frame using the `data` argument

```{r ggplot1, eval=FALSE}
ggplot(data = iprg)
```

* define aesthetics (`aes`), by selecting the variables to be plotted
  and the variables to define the presentation such as plotting size,
  shape color, etc.

```{r ggplot2, eval=FALSE}
ggplot(data = iprg, aes(x = Run, y = Log2Intensity))
```

* add `geoms` -- graphical representation of the data in the plot
	 (points, lines, bars). To add a geom to the plot use `+` operator

```{r ggplot3}
ggplot(data = iprg, aes(x = Run, y = Log2Intensity)) +
  geom_boxplot()
```

See the [documentation page](http://ggplot2.tidyverse.org/reference/)
to explore the many available `geoms`.


The `+` in the `ggplot2` package is particularly useful because it
allows you to modify existing `ggplot` objects. This means you can
easily set up plot "templates" and conveniently explore different
types of plots, so the above plot can also be generated with code like
this:

```{r ggplot4, eval=FALSE}
## Assign plot to a variable
ints_plot <- ggplot(data = iprg, aes(x = Run, y = Log2Intensity))

## Draw the plot
ints_plot + geom_boxplot()
```

Notes:

* Anything you put in the `ggplot()` function can be seen by any geom layers
  that you add (i.e., these are universal plot settings). This includes the x and
  y axis you set up in `aes()`.

* You can also specify aesthetics for a given geom independently of the
  aesthetics defined globally in the `ggplot()` function.

* The `+` sign used to add layers must be placed at the end of each
  line containing a layer. If, instead, the `+` sign is added in the
  line before the other layer, `ggplot2` will not add the new layer
  and will return an error message.


> **Challenge**
>
> * Repeat the plot above but displaying the raw intensities.
> * Log-10 transform the raw intensities on the flight when plotting.

<details>
```{r ggplotchallenge1}
ggplot(data = iprg, aes(x = Run, y = Intensity)) + geom_boxplot()
ggplot(data = iprg, aes(x = Run, y = log10(Intensity))) + geom_boxplot()
```
</details>


## Building complex plots

First, let's colour the boxplot based on the condition:

```{r ggplotbuild1}
ggplot(data = iprg,
	   aes(x = Run, y = Log2Intensity,
		   fill = Condition)) +
  geom_boxplot()
```

Now let's rename all axis labels and title, and rotate the x-axis
labels 90 degrees. We can add those specifications using the `labs`
and `theme` functions of the `ggplot2` package.

```{r ggplotbuild2}
ggplot(aes(x = Run, y = Log2Intensity, fill = Condition),
	   data = iprg) +
	geom_boxplot() +
	labs(title = 'Log2 transformed intensity distribution per MS run',
		 y = 'Log2(Intensity)',
		 x = 'MS run') +
	theme(axis.text.x = element_text(angle = 90))
```


And easily switch from a boxplot to a violin plot representation by
changing the `geom` type.

```{r ggplotbuild3}
ggplot(aes(x = Run, y = Log2Intensity, fill = Condition),
	   data = iprg) +
	geom_violin() +
	labs(title = 'Log2 transformed intensity distribution per Subject',
		 y = 'Log2(Intensity)',
		 x = 'MS run') +
	theme(axis.text.x = element_text(angle = 90))
```

Finally, we can also overlay multiple geoms by simply *adding* them
one after the other.

```{r ggplotbuild4}
p <- ggplot(aes(x = Run, y = Log2Intensity, fill = Condition),
			data = iprg)
p + geom_boxplot()
p + geom_boxplot() + geom_jitter() ## not very usefull
p + geom_jitter() + geom_boxplot()
p + geom_jitter(alpha = 0.1) + geom_boxplot()
```

> **Challenge**
>
> * Overlay a boxplot goem on top of a jitter geom for the raw or
>   log-10 transformed intensities.
> * Customise the plot as suggested above.

<details>
```{r ggplotchallenge2}
## Note how the log10 transformation is applied to both geoms
ggplot(data = iprg, aes(x = Run, y = log10(Intensity))) +
	geom_jitter(alpha = 0.1) +
	geom_boxplot()
```
</details>

Finally, a very useful feature of `ggplot2` is **facetting**, that
defines how to subset the data into different *panels* (facets).

```{r ggplotfacet}
names(iprg)
ggplot(data = iprg,
	   aes(x = TechReplicate, y = Log2Intensity,
		   fill = Condition)) +
	geom_boxplot() +
	facet_grid(~ Condition)
```

## Customising plots

Using `labs`:

```{r labs, eval=FALSE}
p <- ggplot(data = iprg, aes(x = Run, y = log10(Intensity))) +
	geom_jitter(alpha = 0.1) +
	geom_boxplot()
p + labs(title = "A title, at the top",
		 subtitle = "A subtitle, under the title",
		 caption = "Comes at the bottom of the plot",
		 x = "x axis label",
		 y = "y axis label")
p
p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Setting themes:

```{r theme, eval = TRUE}
p1 <- p + theme_bw()
p2 <- p + theme_gray()
p3 <- p + theme_dark()
p4 <- p + theme_minimal()
p5 <- p + theme_light()
p6 <- p + theme_void()
## ....
```

See also the `r CRANpkg("ggthemes")` package.

## Combine `ggplot` figures

> The goal of `r Githubpkg("thomasp85/patchwork")` patchwork is to
> make it ridiculously simple to combine separate ggplots into the
> same graphic. As such it tries to solve the same problem as
> `gridExtra::grid.arrange()` and `cowplot::plot_grid` but using an
> API that incites exploration and iteration.

Installation:

```{r installpatchwork, eval=FALSE}
# install.packages("devtools")
devtools::install_github("thomasp85/patchwork")
```

```{r usepatckwork1, }
library("patchwork")
p1 + p2
p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
p1 + p2 + p3 + plot_layout(ncol = 1)
p1 + p2 - p3 + plot_layout(ncol = 1)
(p1 | p2 | p3) / p4
(p1 + (p2 + p3) + p4 + plot_layout(ncol = 1))
(p1 + (p2 + p3) + p4 + plot_layout(ncol = 1)) * theme_dark()
(p1 + (p2 + p3) + p4 + plot_layout(ncol = 1)) & theme_dark()
```

## Saving your figures

You can save plots to a number of different file formats. PDF is by
far the most common format because it's lightweight, cross-platform
and scales up well but jpegs, pngs and a number of other file formats
are also supported. Let's redo the last barplot but save it to the
file system this time.

Let's save the boxplot as pdf file.

```{r figsave, eval=FALSE}
pdf()
p + geom_jitter(alpha = 0.1) + geom_boxplot()
dev.off()
```

The default file name is `Rplots.pdf`. We can customise that file name
specifying it by passing the file name, as a character, to the `pdf()`
function.

> **Challenge**
>
> Save a figure of your choice to a pdf file. Read the manual for the
> `png` function and save that same image to a png file.
>
> **Tip**: save your figures in a dedicated directory.

## References

- [`ggplot2` extensions - gallery](http://www.ggplot2-exts.org/gallery/)
- [`ggplot2` webpage](http://ggplot2.tidyverse.org/) and documentation
- *ggplot2: Elegant Graphics for Data Analysis* book (source of the
  book available for free [here](https://github.com/hadley/ggplot2-book)).
