# Plotting in R

## Base graphics



## Plotting with `ggplot2`

Base graphics uses a *canvas model* a series of instructions that
sequentially fill the plotting canvas. While this model is very useful
to build plots bits by bits bottom up, which is useful in some cases,
it has some clear drawback:

* Layout choices have to be made without global overview over what may
  still be coming.
* Different functions for different plot types with different
  interfaces.
* No standard data input.
* Many routine tasks require a lot of boilerplate code.
* No concept of facets/lattices/viewports.
* Poor default colours.

The `ggplot2` package implements a grammar of graphics. Users describe
what and how to visualise data and the package then generates the
figure. The components of `ggplot2`'s of graphics are

1. A **tidy** dataset
2. A choice of geometric objects that servers as the visual
   representation of the data - for instance, points, lines,
   rectangles, contours.
3. A description of how the variables in the data are mapped to visual
   properties (aesthetics) or the geometric objects, and an associated
   scale (e.g. linear, logarithmic, polar)
4. A statistical summarisation rule
5. A coordinate system.
6. A facet specification, i.e. the use of several plots to look at the
   same data.



Fist of all, we need to load the `ggplot2` package and load the `iprg` data.

```{r}
library("ggplot2")
iprg <- read.csv("http://bit.ly/VisBiomedDataIprgCsv")
```

ggplot graphics are built step by step by adding new elements.

To build a ggplot we need to:

* bind the plot to a specific data frame using the `data` argument

```{r, eval=FALSE}
ggplot(data = iprg)
```

* define aesthetics (`aes`), by selecting the variables to be plotted
  and the variables to define the presentation such as plotting size,
  shape color, etc.

```{r, eval=FALSE}
ggplot(data = iprg, aes(x = Run, y = Log2Intensity))
```

* add `geoms` -- graphical representation of the data in the plot
	 (points, lines, bars). To add a geom to the plot use `+` operator

```{r}
ggplot(data = iprg, aes(x = Run, y = Log2Intensity)) +
  geom_boxplot()
```

See the [documentation page](http://ggplot2.tidyverse.org/reference/)
to explore the many available `geoms`.


The `+` in the `ggplot2` package is particularly useful because it
allows you to modify existing `ggplot` objects. This means you can
easily set up plot "templates" and conveniently explore different
types of plots, so the above plot can also be generated with code like
this:

```{r, eval=FALSE}
## Assign plot to a variable
ints_plot <- ggplot(data = iprg, aes(x = Run, y = Log2Intensity))

## Draw the plot
ints_plot + geom_boxplot()
```

Notes:

* Anything you put in the `ggplot()` function can be seen by any geom layers
  that you add (i.e., these are universal plot settings). This includes the x and
  y axis you set up in `aes()`.

* You can also specify aesthetics for a given geom independently of the
  aesthetics defined globally in the `ggplot()` function.

* The `+` sign used to add layers must be placed at the end of each
  line containing a layer. If, instead, the `+` sign is added in the
  line before the other layer, `ggplot2` will not add the new layer
  and will return an error message.




## References

- [`ggplot2` extensions - gallery](http://www.ggplot2-exts.org/gallery/)
- [`ggplot2` webpage](http://ggplot2.tidyverse.org/) and documentation
- *ggplot2: Elegant Graphics for Data Analysis* book (source of the
  book available for free [here](t
  https://github.com/hadley/ggplot2-book)).
