# Interactive visualisation

## Interactivity graphs with `ggiz`

> This section is based on the on-line [`ggvis` documentation](http://ggvis.rstudio.com/)

The goal of ggvis is to make it easy to build interactive graphics for
exploratory data analysis. [`ggvis`](http://ggvis.rstudio.com/) has a
similar underlying theory to `ggplot2` (the grammar of graphics), but
it's expressed a little differently, and adds new features to make
your plots interactive. [`ggvis`](http://ggvis.rstudio.com/) also
incorporates shiny's reactive programming model (see later) and
`dplyr`'s grammar of data transformation.

```{r ggvis1, eval = FALSE}
crcdf <- readr::read_csv("./data/CRC_train.csv")
library("ggvis")
```

```{r ggvis2, eval = FALSE}
p <- ggvis(crcdf, x = ~A1AG2, y = ~AFM)
p %>% layer_points()
```

```{r, eval = FALSE}
crcdf %>%
	ggvis(x = ~A1AG2, y = ~AFM,
		  fill = ~Group) %>%
	layer_points()
```

```{r ggvis3, eval = FALSE}
p %>% layer_points(fill = ~Group)
p %>% layer_points(shape = ~Group)
```

To set fixed plotting parameters, use `:=`.

```{r ggvis4, eval = FALSE}
p %>% layer_points(fill := "red", stroke := "black")
p %>% layer_points(size := 300, opacity := 0.4)
p %>% layer_points(shape := "cross")
```

### Interactivity {-}

```{r ggvis5, eval = FALSE}
p %>% layer_points(
		  size := input_slider(10, 200),
		  opacity := input_slider(0, 1))
```
* `add_tooltip`
* `input_slider()`
* `input_checkbox()`
* `input_checkboxgroup()`
* `input_numeric()`
* `input_radiobuttons()`
* `input_select()`
* `input_text()`

See the
[interactivity vignette](http://ggvis.rstudio.com/interactivity.html)
for details.


### Layers {-}

**Simple layers**

* `layer_points()`, with properties x, y, shape, stroke, fill,
  strokeOpacity, fillOpacity, and opacity.
* `layer_paths()`, for paths and polygons (using the fill argument).
* `layer_ribbons()` for filled areas.
* `layer_rects()`, `layer_text()`.

**Compound layers**, which  combine data transformations with one
or more simple layers.

* `layer_lines()` which automatically orders by the x variable with
  `arrange()`.
* `layer_histograms()` and `layer_freqpolys()`, which first bin the
  data with `compute_bin()`.
* `layer_smooths()`, which fits and plots a smooth model to the data
  using `compute_smooth()`.

See the [layers vignette](http://ggvis.rstudio.com/layers.html) for
details.


Like for `ggplot2`'s geoms, we can overly multiple layers:

```{r ggvis7, eval = FALSE}
p %>%
	layer_points() %>%
	layer_smooths(stroke := "red")
```

### More components{-}

* `scales`, to control the mapping between data and visual properties;
  see the
  [properties and scales vignette](http://ggvis.rstudio.com/properties-scales.html).

* `legends` and `axes` to control the appearance of the guides
  produced by the scales. See the
  [axes and legends vignette](http://ggvis.rstudio.com/axes-legends.html).


> Challenge
>
> Apply a PCA analysis on the `mulvey2015norm` data, and use `ggvis` to
> visualise that data along PC1 and PC2, controlling the point size
> using a slide bar.

<details>
```{r ggpca, eval=FALSE}
p <- ggvis(data.frame(pca$x), x = ~PC1, y = ~PC2)
p %>% layer_points(
		  size := input_slider(10, 200),
		  opacity := input_slider(0, 1))
```
</details>


## Interactive apps with `plotly`

The [`plotly`](https://github.com/ropensci/plotly) package can be used
for creating interactive web graphics via the open source JavaScript
graphing library [plotly.js](https://github.com/plotly/plotly.js).

```{r plotly1}
library("plotly")
p <- ggplot(data = crcdf,
			aes(x = A1AG2, y = AFM, colour = Group)) +
	geom_point()
ggplotly(p)
```

```{r plotly2}
crcdf <- crcdf[c(1:10, 191:200), c(70:74)]
x <- gather(crcdf,
			key = Protein, value = x,
			-Sample,
			-Group)
p <- ggplot(data = x,
			aes(x = Sample, y = x,
				group = Protein,
				colour = Protein)) +
		geom_line()


ggplotly(p)
```

See the [plotly for R](https://plotly-book.cpsievert.me/) book for more details.

## Interactive apps

## Motivating examples {-}

In this section, we show how to build interactive `shiny`
applications. `shiny` is widely used to explore and visualise
biomolecular data. As motivating examples, we present here two such
example for proteomics data.

### Longitudinal proteomics QC {-}

`r Biocpkg("MSstatsQC")` uses control charts to monitor the instrument
performance by tracking system suitability metrics including total
peak area, retention time and full width at half maximum (FWHM) and
peak assymetry. Additional metrics can also be analyzed by including
them to the input file.

An online application is available at
https://eralpdogu.shinyapps.io/msstatsqc


### Spatial proteomics {-}

The `r Biocpkg("pRoloc")` package offers dedicated functionality to
manipulate, annotate, analyse and visualise spatial proteomics
data. Its companion package `r Biocpkg("pRolocGUI")` offers support
for interactive visualisation.

You can start an app to visualise one of the published data as shown below

```{r hyperlopit, eval = FALSE}
library("pRoloc")
library("pRolocdata")
data(hyperLOPIT2015)
library("pRolocGUI")
pRolocVis(hyperLOPIT2015)
```

This same data is also available as a standalone on-line
application at https://lgatto.shinyapps.io/christoforou2015

Reference:

- Gatto L, Breckels LM, Wieczorek S, Burger T, Lilley
  KS. Mass-spectrometry-based spatial proteomics data analysis using
  pRoloc and pRolocdata. Bioinformatics. 2014 May 1;30(9):1322-4. doi:
  [10.1093/bioinformatics/btu013](https://doi.org/10.1093/bioinformatics/btu013). Epub
  2014 Jan 11. PubMed PMID: 24413670; PubMed Central PMCID:
  PMC3998135.

- Breckels LM, Mulvey CM, Lilley KS and Gatto L. A Bioconductor
  workflow for processing and analysing spatial proteomics data
  [version 1; referees: 1 approved, 1 approved with
  reservations]. F1000Research 2016, 5:2926 (doi:
  [10.12688/f1000research.10411.1](https://f1000research.com/articles/5-2926/))

### Introduction {-}

A useful `shiny` cheet sheet is available
[here](https://github.com/rstudio/cheatsheets/raw/master/shiny.pdf).

> This section is based on RStudio `shiny` tutorials.

From the [`shiny`](https://shiny.rstudio.com/)
package [website](https://shiny.rstudio.com/):

> Shiny is an R package that makes it easy to build interactive web
> apps straight from R.

When using [`shiny`](https://shiny.rstudio.com/), one tends to aim for
more complete, long-lasting applications, rather then *simply* and
transient visualisations.

A [`shiny`](https://shiny.rstudio.com/) application is composed of a
*ui* (user interface) and a *server* that exchange information using a
programming paradigm called *reactive programming*: changes performed
by the user to the *ui* trigger a reaction by the *server* and the
output is updated accordingly.

- In the ui: define the components of the user interface (such as page
  layout, page title, input options and outputs), i.e what the user
  will see and interact with.

- In the server: defines the computations in the R backend.

- The *reactive* programming is implemented through *reactive*
  functions, which are functions that are only called when their
  respective inputs are changed.

- An application is run with the `shiny::runApp()` function, that
  takes the directory containing the two files as input.

Before looking at the details of such an architecture, let's build a
simple example from scratch, step by step. This app, shown below, uses
the `faithful` data, describing the wainting time between eruptions
and the duration of the reuption for the Old Faithful geyser in
Yellowstone National Park, Wyoming, USA.

```{r faithfuldata}
head(faithful)
```

It shows the distribution of waiting times along a histogram (produced
by the `hist` function) and provides a slider to adjust the number of
bins (the `breaks` argument to `hist`).

```{r shinyapp1, echo=FALSE}
knitr::include_app("https://lgatto.shinyapps.io/shiny-app1/",
				   height = "650px")
```

The app can also be opened at https://lgatto.shinyapps.io/shiny-app1/


### Creation of our fist shiny app {-}

1. Create a directory that will contain the app, such as for example
   `"shinyapp"`.
2. In this directory, create the *ui* and *server* files, named `ui.R`
   and `server.R`.
3. In the `ui.R` file, let's defines a (fluid) page containing
   - a title panel with a page title;
   - a layout containing a sidebar and a main panel

```{r ui00, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app00/ui.R'))
```

4. In the `server.R` file, we define the `shinyServer` function that
   handles input and ouputs (none at this stage) and the R logic.

```{r server00, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app00/server.R'))
```

5. Let's now add some items to the *ui*: a text input widget in the
   sidebar and a field to hold the text ouput.

```{r ui01, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app01/ui.R'))
```

6. In the `server.R` file, we add in the `shinyServer` function some R
   code defining how to manipulate the user-provided text and render
   it using a shiny `textOuput`.

```{r server01, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app01/server.R'))
```

7. Let's now add a plot in the main panel in `ui.R` and some code to
   draw a histogram in `server.R`:

```{r ui02, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app02/ui.R'))
```

```{r server02, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app02/server.R'))
```

8. We want to be able to control the number of breaks used to plot the
   histograms. We first add a `sliderInput` to the *ui* for the user
   to specify the number of bins, and then make use of that new input
   to parametrise the histogram.

```{r ui03, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app03/ui.R'))
```

```{r server03, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app03/server.R'))
```

9. The next addition is to add a menu for the user to choose a set of
   predefined colours (that would be a `selectInput`) in the `ui.R`
   file and use that new input to parametrise the colour of the
   histogramme in the `server.R` file.

```{r ui04, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app04/ui.R'))
```

```{r server04, comment = "", echo=FALSE}
writeLines(readLines('src/shiny-app04/server.R'))
```

10. The last addition that we want is to visualise the actual data in
   the main panel. We add a `dataTableOutput` in `ui.R` and generate
   that table in `server.R` using a `renderDataTable` rendering
   function.

```{r ui1, comment = "", echo=FALSE}
lns <- readLines('src/shiny-app1/ui.R')
writeLines(lns[-grep("^\\s*$", lns)])
```

```{r server1, comment = "", echo=FALSE}
lns <- readLines('src/shiny-app1/server.R')
writeLines(lns[-grep("^\\s*$", lns)])
```

> Challenge
>
> Write and your `shinyapp` applications, as described above.


### Single-file app {-}

Instead of defining the *ui* and *server* in their respective files,
they can be combined into list to be passed directly to `runApp`:

```{r singlefileapp, eval = FALSE}
ui <- fluidPage(...)
server <- function(input, output) { ... }

app <- list(ui = ui, server = server)
runApp(app)
```

> Challenges
>
> 1. Create an app to visualise the `mulvey2015norm` data where the
>    user can select along which features to view the data.
>
> 2. As above, where the visualisation is a PCA plot and the user chooses
>    the PCs.

### There's more to `shiny` {-}

#### Sharing `shiny` apps {-}

* Share the code file(s) and `runApp`
* `runUrl`
* `runGitHub`
* `runGist`
* [shinyapps](http://wwwshinyapps.io)
* Shiny server (in-house)

#### More interactivity {-}

```{r moreinter, eval = FALSE}
 plotOutput("pca",
			hover = "hover",
			click = "click",
			dblclick = "dblClick",
			brush = brushOpts(
				id = "brush",
				resetOnNew = TRUE))
```

Example [here](http://shiny.rstudio.com/gallery/plot-interaction-advanced.html).

#### References {-}

* [`shiny` page](http://shiny.rstudio.com/)
* [`shiny` cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/shiny.pdf)
