---
title: "Visualisation of biomolecular data"
author: "Laurent Gatto"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
description: "Visualisation of biomolecular data"
github-repo: lgatto/VisualisingBiomolecularData
---

```{r setup, echo=FALSE, warning=FALSE}
source("./src/setup.R")
```

# Introduction

This *Visualisation of biomolecular data* course is aimed at people
who are already familiar with the R language and syntax, and who would
like to get a hands-on introduction to visualisation with a focus on
biomolecular data.

## Installation

A set of packages that are used, either directly or indirectly are
provided below. A complete session information with all packages used
to compile this document is available at the end.

The source code for this document is available on GitHub at
https://github.com/lgatto/VisualisingBiomolecularData

A short URL for this book is http://bit.ly/biomolvis


```{r pkgs, echo=FALSE}
library("ggplot2")
library("MSnbase")
library("pRolocdata")
library("pRoloc")
library("ALL")
library("MSstatsBioData")
```

## Questions

For any questions or suggestions, please open an
[issue](https://github.com/lgatto/VisualisingBiomolecularData/issues). Please
do add the output of your session information and, of possible, a
reproducible example describing your question or suggestion.


## License

This material is licensed under the [Creative Commons
Attribution-ShareAlike 3.0
License](http://creativecommons.org/licenses/by-sa/3.0/). Some content
is inspired by other sources though, see the *Credit* section in the
material.

## References

Gatto L, Breckels LM, Naake T, Gibb S. *Visualization of proteomics
data using R and bioconductor*. Proteomics. 2015
Apr;15(8):1375-89. doi:[10.1002/pmic.201400392](https://doi.org/10.1002/pmic.201400392).
PMID: 25690415; PMCID: PMC4510819.

Key M. *A tutorial in displaying mass spectrometry-based proteomic
data using heat maps*. BMC Bioinformatics. 2012;13 Suppl
16:S10. doi:[10.1186/1471-2105-13-S16-S10](http://dx.doi.org/10.1186/1471-2105-13-S16-S10). Epub
2012 Nov 5. PMID: 23176119; PMCID: PMC3489527.

Gatto L, Christoforou A. *Using R and Bioconductor for proteomics data
analysis*. Biochim Biophys Acta. 2014 Jan;1844(1 Pt
A):42-51. doi:[10.1016/j.bbapap.2013.04.032](http://dx.doi.org/10.1016/j.bbapap.2013.04.032). Epub
2013 May 18. PMID: 23692960.
