---
title: "Visualisation of biomolecular data"
author: "Laurent Gatto"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
description: "Visualisation of biomolecular data"
github-repo: lgatto/VisualisingBiomolecularData
---

```{r setup, echo=FALSE, warning=FALSE}
source("./src/setup.R")
```

# Introduction

This *Visualisation of biomolecular data* course is aimed at people
who are already familiar with the R language and syntax, and who would
like to get a hands-on introduction to visualisation with a focus on
biomolecular data. This course is meant to be mostly hands-on, with an
intuitive understanding of the underlying techniques.

An important aspect of data visualisation is data manipulation,
transformation and the format of the data. These topics will be
introduced and documented throughout the course.

A short URL for this book is http://bit.ly/biomolvis

## Installation

A set of packages that are used, either directly or indirectly are
provided below. A complete session information with all packages used
to compile this document is available at the end.

The source code for this document is available on GitHub at
https://github.com/lgatto/VisualisingBiomolecularData


```{r pkgs, eval=FALSE}
library("ggplot2")
library("MSnbase")
library("pRolocdata")
library("pRoloc")
library("ALL")
library("MSstatsBioData")
```

## Questions

For any questions or suggestions, please open an
[issue](https://github.com/lgatto/VisualisingBiomolecularData/issues). Please
do add the output of your session information and, of possible, a
reproducible example describing your question or suggestion.


## License

This material is licensed under the [Creative Commons
Attribution-ShareAlike 4.0
License](http://creativecommons.org/licenses/by-sa/4.0/). Some content
is inspired by other sources, see the *Credit* section in the
material.

## References

Gatto L, Breckels LM, Naake T, Gibb S. *Visualization of proteomics
data using R and bioconductor*. Proteomics. 2015
Apr;15(8):1375-89. doi: [10.1002/pmic.201400392](https://doi.org/10.1002/pmic.201400392).
PMID: 25690415; PMCID: PMC4510819.

Key M. *A tutorial in displaying mass spectrometry-based proteomic
data using heat maps*. BMC Bioinformatics. 2012;13 Suppl
16:S10. doi: [10.1186/1471-2105-13-S16-S10](http://dx.doi.org/10.1186/1471-2105-13-S16-S10). Epub
2012 Nov 5. PMID: 23176119; PMCID: PMC3489527.

Gatto L, Christoforou A. *Using R and Bioconductor for proteomics data
analysis*. Biochim Biophys Acta. 2014 Jan;1844(1 Pt
A):42-51. doi: [10.1016/j.bbapap.2013.04.032](http://dx.doi.org/10.1016/j.bbapap.2013.04.032). Epub
2013 May 18. PMID: 23692960.
